---
title: Volatility Targeting Using VIX
subtitle:  with applications to portfolio management
author: Jeff Li
date: "`r format(Sys.time(), '%d %B, %Y')`"
fontsize: 12
output:
  beamer_presentation:
    slide_level: 3
    citation_package: natbib
bibliography: references.bib 
---

### Introduction

 - It is well known that volatility clusters in markets tends to "cluster" [@DING1996185].
    That is, current high volatility predicts future high volatility, while
    the same is true for low volatility.
 - Using this empirical fact, one can construct a trading strategy that leverages up 
    during periods of low volatility and leverage down during periods of high volatility.
 - In this presentation, we employ the trading strategy described by [@HARVEY2018].
    - We first replicate their strategy on SPX using VIX as our measure of volatility, 
    we then apply it to other long-only portfolios such as Value and Momentum as described by 
    Kenneth R. French's website [@KENFRENCH].

### S&P 500's Daily Returns From 2000 to 2021

```{r sp500_daily_returns, echo=FALSE}
library(ggplot2)
sp500_excess_returns <- read.csv(
    '/home/jeff/AssetMark/data/spx_return_excess.csv'
    )
sp500_excess_returns$date <- as.Date(sp500_excess_returns$date)
ggplot(data = sp500_excess_returns, aes(date, spx_returns, group = 1)) +
 geom_line() +
 theme_bw() +
 theme(
     panel.border = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     axis.line = element_line(colour = "black")
     )
```

### Strategy Methodology

 - This strategy aims to create a "target" volatility for some particular portfolio.
    - We aim to have a target of 20%, annualized.
 - We leverage up or down according the 2-day lagged VIX values.
 - In essence, if $r_t$ is the excess return\footnote{Excess of risk-free rate.} of our base portfolio at time $t$, then 
    the return of our strategy (volatility targetted), $r_t'$, is 
    $$r_t' = \left(\frac{20\%}{\sigma_{t-2}}\right)r_t k$$
    where $\sigma_t$ is the VIX at time $t$.\footnote{To be precise, it's the VIX index divided by 100.} $k$ is 
    chosen ex-post so our overall backtest will have a historical volatility of 20%.
 - Due to the volatile nature of the VIX, we set caps and floors for our leverage constraints, which is a 50% floor
    and a 200% cap.

### Applying to S&P 500
```{r spx_results, echo=FALSE}
results_table <- read.csv('./data/SPX_results_table.csv')
knitr::kable(results_table)
```

### Total P&L
```{r pnl plot, echo=FALSE, fig.width = 14}
data <- read.csv('./data/pnl.csv')
data$date <- as.Date(data$date)
ggplot(data = data, aes(date, PnL, color = variable, group = 1)) +
 geom_point() +
 theme_bw() +
 theme(
     panel.border = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     axis.line = element_line(colour = "black")
     ) +
 labs(
    caption = "This is different from [Harvey et al., 2018]
    as their data starts from 1935."
    )
```

### Historical Leverage/Weights

```{r strategy weights, echo=FALSE, fig.width = 14}
data <- read.csv('./data/strategy_weights.csv')
data$date <- as.Date(data$date)
ggplot(data = data, aes(date, weights, group = 1)) +
 geom_point() +
 theme_bw() +
 theme(
     panel.border = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     axis.line = element_line(colour = "black")
     )
```

### Rolling 30 Day Vol

```{r vol plot, echo=FALSE, fig.width = 14}
data <- read.csv('./data/30dayvol.csv')
data$date <- as.Date(data$date)
ggplot(data = data, aes(date, rolling30DayVol, color = variable, group = 1)) +
 geom_point() +
 theme_bw() +
 theme(
     panel.border = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     axis.line = element_line(colour = "black")
     )
```

### 1% Left Tail of Distribution
```{r shortfall plot, echo=FALSE, fig.width = 14}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
data <- read.csv('./data/shortfall_returns.csv')
data_spx <- na.omit(data$SPX)
data_spx_target <- na.omit(data$SPX_target)
d <- data.frame(x = c(data_spx, data_spx_target),
            type = rep(c("SPX 1% Left Tail", "SPX Targetted Vol 1% Left Tail"),
            c(length(data_spx), length(data_spx_target))))
ggplot(d) +
  geom_density(aes(x = x, colour = type, fill = type), alpha = 0.5) +
  theme_bw() +
   theme(
     panel.border = element_blank(),
     panel.grid.major = element_blank(),
     panel.grid.minor = element_blank(),
     axis.line = element_line(colour = "black")
     ) + 
   xlab("Daily Returns") +
   ylab("Probability")

```

### Applying to Top Decile of a Momentum Sort
```{r momentum results, echo=FALSE}
results_table <- read.csv('./data/momentum_results_table.csv')
knitr::kable(results_table)
```


### Applying to Top Decile of a B/M Sort
```{r value results, echo=FALSE}
results_table <- read.csv('./data/value_results_table.csv')
knitr::kable(results_table)
```

### Discussion
 - We see that using VIX to create volatility targetting strategies
   seem to improve on long-only strategies.

 - However, these improvements comes at a significant increase in turnover.

 - It could be the case that the increase in trading costs is the source of 
   additional returns.

 - In terms of implementing these strategies, doing it for S&P 500 related 
   assets is the most feasiable due to liquidity constraints.
      - One can trade S&P 500 futures while Momentum and Value portfolios
         would require trading a lot of stocks.

 - Compare the difference in performance of VIX driven volatility targetting
   compared to standard deviation driven volatility targetting